Analysis of observational data, to calculate the Delayed-Tau model parameters

### 2025-03-26 thesis.qmd
final results+conclusion

  @haslbauerCosmologicalStarFormation2023 argue that the Delayed-$\tau$ model is opposed to the Lilly-Madau plot ([@madauCosmicStarFormation2014]), which plots the observed SFR's of galaxies with the corresponding redshifts ($z$) and calculates a cosmic SFR peak at $z\approx 2$. The way they calculated this inconsistency is by using observatory data for SFR and Stellar Masses from the UNGC catalog (@karachentsevUPDATEDNEARBYGALAXY2013, @karachentsevSTARFORMATIONPROPERTIES2013) for calculating the parameters (the timescale $\tau$ and the normalization constant $A_{del}$) of the model. This calculation for the galaxies of the Local Cosmological Volume (LV), allows the investigation of the SFR throughout the life of each galaxy and so we can find the expected time of peak of the SFR.

@@

where $\tau$ is the timescale, $t_{\text{sf}} = t - T_0$ is the star formation time, $t$ is the age of the Universe, and $T_0$ is the time when star formation began.

@@

$$
\text{SFR}_0 = \text{SFR}(t_{sf}) = A_{del}\frac{t_{sf}}{\tau^2}e^{-t_{sf}/\tau}
$$ {#eq-delayed-current}

@@

The specific-SFR (sSFR) describes the rate of change of the stellar mass in a galaxy. Assuming a function $g(x)$ and its derivative $g'(x)$ for the Newton-Raphson algorithm:

@@

![Paramater distribution of the delayed-$\tau$ SFH for the galaxies of the LCV. The colourbar represents the present-day stellar mass of the galaxies. The left plot shows the whole parameter range of the model, including negative timescales and $A_{del}$ values that occur by instabilities. In the right plot, the region $\tau \in [0,16]$ and $A_{del}\in[0,10^{13}]$ is depicted. The majority of the galaxies have a timescale of about 3.5-4.5 Gyr as expected from @speagleHighlyConsistentFramework2014](NR/tau_A_double_plot.png){#fig-A-tau}

@@

![$A_{del}-\text{sSFR}$ plot, where $A_{del}$ is calculated with different methods. In blue and green are the normalization constant values as they calculated by the MCMC method with different priors, while in red are results from Newton-Raphson.](method_comparison/sSFR_A.png){#fig-sSFR-A}

@@

we can create a MCMC model to calculate the parameters, with the priors:

-   $t_{sf}\sim \text{uniform(lower bound = 1 Gyr, upper bound = 13.8 Gyr)}$,
-   $\zeta \sim \text{normal(mean=1.3, }\sigma = 0.01)$, within the limits $[1,2]$,

and the likelihood $\log(\text{SFR}_{obs}) \sim \text{normal}\left(\log(\text{SFR}_\text{MCMC}), \sigma = 0.1\right)$, while taking the Stellar Masses of the galaxies as they are, without any variance, and $A_{del}$ is calculated by @eq-A. For the $\tau$ we will consider two different priors, $\tau \sim \text{uniform(lower bound = 1 Gyr, upper bound = 13.8 Gyr)}$ and $\tau \sim \text{normal(mean = 4 Gyr, }\sigma  = 13.8 \text{Gyr)}$. The MCMC run consists of 6 chains with 8000 iterations each (2500 iterations for the burn-in phase). Using this method, the galaxies can have different $t_{sf}$, while only taking positive values for their timescales.

The prior of the $\tau$ and $t_{sf}$ follow a uniform distribution, since we do not want to introduce any unwanted bias in the value selection of the MCMC sampler. However, there is bibliography that could justify using a normal distribution for $\tau$, for example @speagleHighlyConsistentFramework2014 and @kroupaConstraintsStarFormation2020 explain that the galaxies of the Main Sequence are expected to have $3.5\lesssim\tau/\text{Gyr}\lesssim4.5$. The same prior will not apply for $t_{sf}$ since it could significatly effect the results, especially of younger galaxies.

Using these priors, the metrics of the Stan programm give good results ($\hat{R}<1.001$ and $\text{ESS} >1000$), while the calculated values correspond to the data of the catalog. Specifically, we use the data to recalculate the SFR from the predicted parameters, using the delayed-$\tau$ model and compare the normalization constant to the stellar masses of the galaxies (@fig-A-M-mcmc).

::: {#fig-A-M-mcmc layout="[[50,50],[70]]"}
![Uniform $\tau$ prior](method_comparison/A_M_MCMC_uni.png){#fig-A-M-mcmc-uni}

@@

![Newton-Raphson Method](/NR/A-M_*-c_x.png){#fig-A-M}

Linear regression for $A_{del}-M_*$ for the $A_{del}$ values from the two MCMC runs and the Newton-Raphson method, with the colorbar repressenting the ratio $x=t_{sf}/\tau$. For the MCMC runs, the black dashed line represents the regresion weighted by $x$. For the upper left plot, the regression yields a correlation of 94%, while the weighted regression is the fit $\log(A_{del}) = 0.95\cdot\log(M_*)+0.60$, with a correlation of 98%. For the upper right plot, the correlation is higher, at 99%, with a weighted regression of $\log(A_{del}) = 0.97\cdot\log(M_*)+0.44$, with a correlation of 99.5%. For the N-R plot, we do two different regressions for the two distinct regions of $A_{del}$, with the upper region only having a correlation of 55\%, while the lower region closely relating the normalization constant with the Stellar Masses of the galaxies, as described in the MCMC plots.
:::

In @fig-sSFR-A, it can be seen that the MCMC run does not have the same inconsistancies with the NR method, and @fig-tau-hist-both shows that the MCMC method captures better both the galaxies with small timescales, and the galaxies with slow rising SFR's.

::: {#fig-distr-comparison layout-ncol="2"}
![Distribution of $\tau$ values.](method_comparison/tau_hist.png){#fig-tau-hist-both}

Distribution of the results of $\tau$ and $t_{sf}$ from all the methods. In blue are the distributions as they are calculated by the MCMC method with the uniform $\tau$ prior, in green are the results from the MCMC run with the normal prior, while in red are results from Newton-Raphson.
:::

@@

Having calculated the model parameters, it is possible to recunstruct the SFH of the galaxies, and from the @eq-SFRD, it is possible to calculate the SFRD of the LCV and recreate the Lilly-Madau Plot. As seen from @fig-lilly-madau, the results from the MCMC method are more consistent with the theoretical line give by the [@eq-lilly-madau; @madauCosmicStarFormation2014], while the NR method underastimates the SFRD of the LCV, by a factor[^11] greater than $\sim 3$, up until $z \approx 3$. From the three methods, the MCMC with the prior $\tau\sim \text{normal}$ is systematicaly closer to universal SFRD, indicating that the assumption for the prior is better than the uniform prior, while the NR line can not approximate the expected line. Eventhough the MCMC method could approximate the SFRD for small redshifts, at greater $z$ the model can overestimate the SFRD by a factor of $\sim 2-3.2$.

[^11]: The factor is calculated by the residuals, $\log(SFRD_\text{data})-\log(SFRD_{LM})=\log\left(\frac{SFRD_\text{data}}{SFRD_{LM}}\right)\Leftrightarrow \frac{SFRD_\text{data}}{SFRD_{LM}} = 10^{\log(SFRD_\text{data})-\log(SFRD_{LM})}$

::: {#fig-lilly-madau layout-ncol="2"}
![SFH Reconstruction of the Local Cosmological Volume](method_comparison/sfrd_comparison_custom_axes_log_values.png)

![Residuals of the SFH reconstruction of the LCV based on the theoritical line given by @madauCosmicStarFormation2014](method_comparison/sfrd_residuals_vs_lm.png)

@@

>>>>

  @haslbauerCosmologicalStarFormation2023 argue that the use of Delayed-$\tau$ model for the galaxies of the Local Cosmological Volume (LCV) is opposed to the Lilly-Madau plot ([@madauCosmicStarFormation2014]), which plots the observed SFR's of galaxies with the corresponding redshifts ($z$) and calculates a cosmic SFR peak at $z\approx 2$. The way they calculated this inconsistency is by using observatory data for SFR and Stellar Masses from the UNGC catalog (@karachentsevUPDATEDNEARBYGALAXY2013, @karachentsevSTARFORMATIONPROPERTIES2013) for calculating the parameters (the timescale $\tau$ and the normalization constant $A_{del}$) of the model. This calculation for the galaxies of the Local Cosmological Volume (LV), allows the investigation of the SFR throughout the life of each galaxy and so we can find the expected time of peak of the SFR.

@@

where $\tau$ is the timescale, $t_{\text{sf}} = T_0 - t_\text{start}$ is the star formation time, $T_0$ is the age of the Universe, and $t_\text{start}$ is the time when star formation began.

@@

$$
\text{SFR}_0 = \text{SFR}(t_{sf}) = A_{del}\frac{t_{sf}}{\tau^2}e^{-t_{sf}/\tau},\ t_{sf} = T_0-t_\text{start}
$$ {#eq-delayed-current}

@@

The specific-SFR (sSFR) provides a normalized view of star formation intensity. Assuming a function $g(x)$ and its derivative $g'(x)$ for the Newton-Raphson algorithm:

@@

![Paramater distribution of the delayed-$\tau$ SFH for the galaxies of the LCV. The colorbar represents the present-day stellar mass of the galaxies. The left plot shows the whole parameter range of the model, including negative timescales and $A_{del}$ values that occur by instabilities. In the right plot, the region $\tau \in [0,16]$ and $A_{del}\in[0,10^{13}]$ is depicted. The majority of the galaxies have a timescale of about 3.5-4.5 Gyr as expected from @speagleHighlyConsistentFramework2014](NR/tau_A_double_plot.png){#fig-A-tau}

@@

![$A_{del}-\text{sSFR}$ plot, where $A_{del}$ is calculated with different methods.](method_comparison/sSFR_A.png){#fig-sSFR-A}

@@

we can create a MCMC model to calculate the parameters, with the priors $t_{sf}\sim \text{uniform(lower bound = 1 Gyr, upper bound = 13.8 Gyr)}$, $\zeta \sim \text{normal(mean=1.3, }\sigma = 0.01)$, within the limits $[1,2]$ and the likelihood $\log(\text{SFR}_{obs}) \sim \text{normal}\left(\log(\text{SFR}_\text{MCMC}), \sigma = 0.1\right)$, while taking the Stellar Masses of the galaxies as they are, without any variance, and $A_{del}$ is calculated by @eq-A. For the $\tau$ we will consider two different priors, $\tau \sim \text{uniform(lower bound = 1 Gyr, upper bound = 13.8 Gyr)}$ and $\tau \sim \text{normal(mean = 4 Gyr, }\sigma  = 13.8 \text{Gyr)}$. The MCMC run consists of 6 chains with 8000 iterations each (2500 iterations for the burn-in phase). Using this method, the galaxies can have different $t_{sf}$, while only taking positive values for their timescales.

The prior of the $\tau$ and $t_{sf}$ follow a uniform distribution, since we do not want to introduce any unwanted bias in the value selection of the MCMC sampler. However, there is bibliography that could justify using a normal distribution for $\tau$, for example @speagleHighlyConsistentFramework2014 and @kroupaConstraintsStarFormation2020 explain that the galaxies of the Main Sequence are expected to have $3.5\lesssim\tau/\text{Gyr}\lesssim4.5$. The same prior will not apply for $t_{sf}$ since it could significantly effect the results, especially of younger galaxies.

Using these priors, the metrics of the Stan program give good results ($\hat{R}<1.001$ and $\text{ESS} >1000$), while the calculated values correspond to the data of the catalog. Specifically, we use the data to recalculate the SFR from the predicted parameters, using the delayed-$\tau$ model and compare the normalization constant to the stellar masses of the galaxies (@fig-A-M-mcmc).

::: {#fig-distr-comparison layout-ncol="2" fig-pos="b"}
![Distribution of $\tau$ values.](method_comparison/tau_hist.png){#fig-tau-hist-both}

Distribution of the results of $\tau$ and $t_{sf}$ from all the methods.
:::

In @fig-sSFR-A, it can be seen that the MCMC run does not have the same inconsistencies with the N-R method, and @fig-tau-hist-both shows that the MCMC method captures better both the galaxies with small timescales, and the galaxies with slow rising SFR's.

::: {#fig-A-M-mcmc layout="[[50,50],[1]]" fig-pos="hb"}
![Uniform $\tau$ prior](method_comparison/A_M_MCMC_uni.png){#fig-A-M-mcmc-uni}

@@

![Newton-Raphson Method](/NR/A-M_*-c_x.png){#fig-A-M width="60%"}

Linear regression for $A_{del}-M_*$ for the $A_{del}$ values from the two MCMC runs and the Newton-Raphson method, with the colorbar representing the ratio $x=t_{sf}/\tau$. For the MCMC runs, the black dashed line represents the regression weighted by $x$. For the upper left plot, the regression yields a correlation of 94%, while the weighted regression is the fit $\log(A_{del}) = 0.95\cdot\log(M_*)+0.60$, with a correlation of 98%. For the upper right plot, the correlation is higher, at 99%, with a weighted regression of $\log(A_{del}) = 0.97\cdot\log(M_*)+0.44$, with a correlation of 99.5%. For the N-R plot, we do two different regressions for the two distinct regions of $A_{del}$, with the upper region only having a correlation of 55%, while the lower region closely relating the normalization constant with the Stellar Masses of the galaxies, as described in the MCMC plots.
:::

@@

Having calculated the model parameters, it is possible to reconstruct the SFH of the galaxies, and from the @eq-SFRD, it is possible to calculate the SFRD of the LCV and recreate the Lilly-Madau Plot. Using the equations @eq-delayed-current and @eq-SFRD, we get the equation:

$$
\text{SFRD(T)} = \frac{1}{V}\sum^N_{i=1} A_{del}\frac{T-t_\text{start}}{\tau^2}\exp\left(-\frac{T-t_\text{start}}{\tau}\right),
$$ {#eq-SFRD-delayed}

where $t_\text{start}$, is calculated by the computed $t_{sf}=T_0-t_\text{start}$, and $T$ is the age of the universe ($T_0$ is the current age=13.8 Gyr). Using this equation we can recreate the parameterized SFRD.

<!-- ![Residuals of the SFH reconstruction of the LCV based on the theoritical line given by @madauCosmicStarFormation2014](method_comparison/sfrd_residuals_vs_lm.png) -->

::: {#fig-lilly-madau-rec layout-ncol="2"}
![SFH Reconstruction of the Local Cosmological Volume](method_comparison/sfrd_comparison_custom_axes_log_values.png)

![Ratio between the cosmic SFRD measured by @madauCosmicStarFormation2014 and the SFRD of the LCV](method_comparison/sfrd_ratio_vs_lm.png)

@@

As seen from @fig-lilly-madau-rec, the MCMC method can approximate better the SFRD of the LCV up until $z\sim 2$ with a factor less than 2. At the peak of the cosmic SFRD, at $z=1.86$, the uniform and normal priors underestimate the SFRD by a factor of 1.82 and 1.53 respectively, while the N-R underestimates it by a factor of 3.6. In the paper, @haslbauerCosmologicalStarFormation2023, they estimate a factor of 2.16 at the peak, using the SFR's from the UNGC catalogs. However, the ratio of the MCMC grows with $z$, since many galaxies start forming stars at a lookback time of $\sim 12$ Gyr, as seen in @fig-tsf-hist-both (the mean value of the $t_{sf}$ of the galaxies is 10 Gyr for the uniform prior and 9.8 Gyr for the normal prior), resulting in extreme deviations from the cosmic SFRD.

# Conclusions

With the use of IR data for the SFR of the LCV and with the use of different algorithms, we were able to reduce the divergence of the SFRD of the LCV from the cosmic SFRD from a factor of 2.16 to 1.53 at $z=1.86$. The N-R, can describe better the SFRD of the LCV compared to the MCMC for higher redshifts, but fails to approximate it for lower redshifts. The high divergence of the MCMC approximations, for $z>2$ is a result of the wrong prior of the $t_{sf}$. Using uniform priors for the time of star formations, severely underestimates the time in which most galaxies start the star forming process, meaning that most of the galaxies stop contributing at the SFRD at earlier lookback times than they should.

A better assumption for the prior of $t_{sf}$, which pushes the mean $\overline{t_{start}}\approx0.2$ Gyr, could correct the model at higher redshifts, since it would approximate the assumption of the N-R, while also keeping the lower factor of the MCMC at the peak of the Lilly-Madau plot. A possible prior for the $t_{sf}$ could be a skewed normal distribution with a mean value $\overline{t_{sf}}\approx 13$ Gyr and with a large sigma to cover the younger galaxies of the LCV. Another possible distribution could be a log-normal distribution for the $t_{start}$. But since the choice of a prior can significantly affect the results and because of the lack of a bibliography for this model, for this undergraduate thesis, the choice of a uniform is the safest choice.

Based on our results, we cannot safely assume that the model works for the galaxies of the LCV. Since parameterized SFH models do not account for bursty star formation, events characteristic of starburst galaxies, incorporating additional parameters who depend on $\text{SFR}_{H\alpha}$ and $\text{SFR}_\text{FUV}$ could increase the accuracy of the model. Additionaly, only examining one model ignores the intrinsic characteristics of the galaxies. @haslbauerCosmologicalStarFormation2023, introduce a combined SFH model depending on the mass of the galaxies, which can reduce the difference of the predicted model and the cosmic SFRD.

In summary, the use of the MCMC and the data from the HECATE catalog proves that the delayed-$\tau$ model a better approximation of the SFH, as originally argued by @haslbauerCosmologicalStarFormation2023. However, more fine-tuning of the model is needed to be certain if the delayed-$\tau$ model is a robust approximation of the SFH of the galaxies of the Local Cosmological Volume.
